Annotated example of a spatially varying effect with sdmTMB

### 2021-11-26 doc/appendix-snow-sdmTMB.Rmd
add refs
edits to owl appendix

---
title: "Annotated example of a spatially varying effect with sdmTMB"
output:
  pdf_document:
    highlight: tango
---

```{r setup, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE
)
```

```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
library(ggplot2)
library(dplyr)
library(sdmTMB)
# library(sf)
theme_set(theme_bw())
# theme_set(theme_minimal())
```

@@

Here, X and Y are coordinates in Albers projection for North America and divided by 100000 to give units of 100 km.

@@

```{r plot-mesh, fig.asp=0.9, out.width="6in"}
plot(mesh)
```

@@

```{r explore, fig.asp=0.4, fig.width = 6}
ggplot(snow, aes(year, count, colour = nao)) + 
  geom_jitter(alpha = 0.5) + 
  scale_colour_viridis_c()
```

@@

```{r load-models, warning=FALSE, message=FALSE, echo=FALSE}
m0 <- readRDS(here::here("owls/snow_w_main_effect_0_150km.rds"))
m1 <- readRDS(here::here("owls/snow_w_main_effect_1_150km.rds"))
m2 <- readRDS(here::here("owls/snow_w_main_effect_150km.rds"))
m <- readRDS(here::here("owls/snow_w_main_effect_150km_reml.rds"))
```

@@

The `nbinom2` distribution appears to have the best fit. 
We will rerun with `reml = TRUE` to ensure better estimation of the random effects that dominate this model and then check our residuals.

@@

```{r print-fit, message=FALSE}
print(m)
```

@@

We can also renerate predictions and residuals for the original data locations:

```{r predict, message=FALSE}
pred <- predict(m)
```

We can inspect randomized quantile residuals, although this method of checking residuals is not guaranteed to be accurate. 
An additional step using STAN could be added.

```{r qqnorm, fig.asp=0.9, out.width="3in"}
pred$resid <- residuals(m)
qqnorm(pred$resid)
qqline(pred$resid)
```

```{r plot-resids, fig.width = 7}
ggplot(pred, aes(X, Y, colour = resid)) +
  geom_point(size = 0.5) +
  facet_wrap(~year) +
  coord_fixed() +
  scale_colour_gradient2()
```

@@

```{r shapes, echo=FALSE, message=FALSE}
library("rnaturalearth")
library("rnaturalearthdata")
coast <- ne_coastline(scale = "medium", returnclass = "sf") %>% 
  sf::st_transform(crs = proj)
lakes <- sf::st_read(here::here("owls/ne_10m_lakes"))
lakes <- lakes[lakes$scalerank==0,] %>% sf::st_transform(crs = proj) # only large lakes
```

@@

```{r plot-map, echo=TRUE, eval=T}
ggplot(data = p_proj) +
  geom_point(data = p_mean, 
     aes(X,Y, colour = nao_effect, size = mean_est_count), alpha = 0.5) +
  geom_sf(data = coast, fill = NA) +
  geom_sf(data = lakes, fill = NA) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),# adjusts space on sides
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(
    direction = "horizontal", title.position = "top", label.position = "bottom")) +
  guides(size = "none") + 
  labs(colour = "NAO effect\non Snowy Owl count") +
  theme_bw() + theme(
    legend.key.height = unit(0.3, "cm"),
    legend.position = c(0.25,0.13),
    axis.title = element_blank())
```

@@

These are the predictions of winter owl counts that incorporate all fixed effects and random effects:

```{r plot-all-effects, echo=TRUE, eval=T, fig.width = 7}
ggplot(data = filter(p_proj, year > 1995)) +
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = exp(est)), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_colour_viridis_c(
    # transform to make variation in low number visible 
    trans="sqrt", 
    # and also trim the top extremes of the colour axis 
    limits = c(0, quantile(exp(p_proj$est), 0.995)), na.value = "yellow") +
  labs(x = "Longitude", y = "Latitude", colour = "Predicted\ncount") +
  theme(axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        axis.title = element_blank()) +
  facet_wrap(~year_f)
```
Note that for these annually varying estimates, we are showing only the most recent 25 years.

## Other components of predicted counts

In addition to the spatially varying effect we started with, there are three other components that go into the overall predictions that we can plot spatially.

1. Just fixed effects

@@

```{r plot-fixed-effects, echo=TRUE, eval=T}
ggplot(data = filter(p_proj, year > 1995)) +
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = exp(est_non_rf)), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_colour_viridis_c() + 
  theme(axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        axis.title = element_blank()) +
  facet_wrap(~year_f)
```

2. Static spatial random effects

@@

```{r plot-spatial-effects, echo=TRUE, eval=T}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X,Y, colour = omega_s), alpha = 0.5) +
  geom_sf(data = coast, fill = NA) +
  geom_sf(data = lakes, fill = NA) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(
    direction = "horizontal", title.position = "top", label.position = "bottom")) +
  theme_bw() + theme(
    legend.key.height = unit(0.3, "cm"),
    legend.position = c(0.25,0.13),
    axis.title = element_blank())
```

3. Spatiotemporal random effects

@@

```{r plot-spatiotemporal-effects, echo=TRUE, eval=T}
ggplot(data = filter(p_proj, year > 1995)) +
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = epsilon_st), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c() +
  theme(axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        axis.title = element_blank()) +
  facet_wrap(~year_f)
```

>>>>

---
title: "Annotated example of a spatially varying effect with sdmTMB"
output:
  pdf_document:
    highlight: tango
bibliography: refs.bib
---

```{r setup, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 1,
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE
)
```

```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
library(ggplot2)
library(dplyr)
library(sdmTMB)
library(sf)
theme_set(theme_bw() + theme(
  legend.title = element_text(size = 10, hjust = 0), legend.key.height = unit(0.3, "cm"),
  legend.position = "top", legend.justification = c(0, 0), 
  axis.title = element_blank()))
```

@@

Here, X and Y are coordinates in Albers projection for North America and divided by 100000 to give units of 100 km. 
This projection was achieved with the following crs code using `sf::st_transform(crs = proj)`.

@@

```{r plot-mesh, out.width="6in"}
plot(mesh)
```

@@

```{r explore, fig.asp = 0.3}
ggplot(snow, aes(year, count, colour = nao)) + 
  geom_jitter() + 
  scale_colour_viridis_c() +
  theme_bw() 
```

@@

```{r load-models, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE}
m0 <- readRDS(here::here("owls/snow_w_main_effect_0_150km.rds"))
m1 <- readRDS(here::here("owls/snow_w_main_effect_1_150km.rds"))
m2 <- readRDS(here::here("owls/snow_w_main_effect_150km.rds"))
m <- readRDS(here::here("owls/snow_w_main_effect_150km_reml.rds"))
```

@@

The `nbinom2` distribution appears to have the best fit. 
We will rerun with `reml = TRUE` to ensure better estimation of the random effects that dominate this model.

@@

```{r print-fit, warning=FALSE, echo=TRUE, message=FALSE}
print(m)
```

@@

We can then generate predictions and residuals for the original data locations:

```{r predict, message=FALSE}
pred <- predict(m)
pred$resid <- residuals(m)
```

We can inspect randomized quantile residuals, although this method of checking residuals is not guaranteed to be accurate.

```{r qqnorm, out.width = "33%"}
qqnorm(pred$resid)
qqline(pred$resid)
```

More accurate residuals can be generated using STAN, but this is a very slow process!
In this case, it took more than 2.5 hrs.

```{r stan-resids, eval=FALSE}
library(tmbstan)
stan_fit <- tmbstan::tmbstan(m$tmb_obj, iter = 100, chains = 1, warmup = 99)
eta <- predict(m, tmbstan_model = stan_fit)
mu <- m$family$linkinv(as.numeric(eta))
r_mcmc <- sdmTMB:::qres_nbinom2(m, y = m$data$count, mu = mu)
saveRDS(r_mcmc, "mcmc_residuals.rds")
```

```{r stan-dat, echo=FALSE, cache=FALSE}
r_mcmc <- readRDS(here::here("owls/mcmc_residuals.rds"))
```

```{r stan-qqnorm, out.width = "33%"}
qqnorm(r_mcmc)
qqline(r_mcmc) 
```

<!-- \clearpage -->

```{r plot-resids, fig.width = 8}
ggplot(pred, aes(X, Y, colour = resid)) +
  geom_point(size = 0.5) +
  facet_wrap(~year) +
  coord_fixed() +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  scale_colour_gradient2(guide = guide_colourbar(title.position = "top"))
```

<!-- \clearpage -->

@@

```{r shapes, warning=FALSE, echo=TRUE, message=FALSE}
library("rnaturalearth")
library("rnaturalearthdata")
coast <- ne_coastline(scale = "medium", returnclass = "sf") %>% 
  sf::st_transform(crs = proj)
lakes <- sf::st_read(here::here("owls/ne_10m_lakes"),  quiet = TRUE)
lakes <- lakes[lakes$scalerank==0,] %>% sf::st_transform(crs = proj) # only large lakes
```

@@

```{r plot-map, echo=TRUE, eval=T}
ggplot(data = p_proj) +
  geom_point(data = p_mean, 
     aes(X,Y, colour = nao_effect, size = mean_est_count), alpha = 0.5) +
  geom_sf(data = coast, fill = NA) +
  geom_sf(data = lakes, fill = NA) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),# adjusts space on sides
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(
    direction = "horizontal", title.position = "top")) +
  guides(size = "none") + 
  labs(colour = "NAO effect\non Snowy Owl count") +
  theme_bw() + theme(
    legend.key.height = unit(0.3, "cm"),
    legend.position = c(0.25,0.13),
    axis.title = element_blank())
```

\clearpage

@@

These are the predictions of winter owl counts that incorporate all fixed effects and random effects. 
The colour axis has been transformed and trimmed to improve the visibility of variation in count estimates.

```{r plot-all-effects, echo=TRUE, eval=T}
ggplot(data = filter(p_proj, year > 1995)) +
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = exp(est)), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_colour_viridis_c(trans="sqrt", # makes variation in low numbers visible 
    # can also trim the extreme high values
    limits = c(0, quantile(exp(p$est), 0.995)), na.value = "yellow",
    guide = guide_colourbar(title.position = "top")) +
  labs(colour = "Predicted counts") + 
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

Note that for all of these annually varying predictions, we are showing only the most recent 25 years.

\clearpage
## Other components of predicted counts

In addition to the spatially varying effect that we started with, there are three other components that go into the overall predictions that we can plot spatially.

### 1. Just fixed effects

@@

```{r plot-fixed-effects, echo=TRUE, eval=T}
ggplot(data = filter(p_proj, year > 1995)) +
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = exp(est_non_rf)), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y)))+
  scale_colour_viridis_c(breaks = c(0.03, 0.05, 0.07), 
    guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

\clearpage

### 2. Static spatial random effects

@@

```{r plot-spatial-effects, echo=TRUE, eval=T}
ggplot(data = p_proj) +
  geom_point(data = p, aes(X,Y, colour = omega_s), alpha = 0.5) +
  geom_sf(data = coast, fill = NA) +
  geom_sf(data = lakes, fill = NA) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000),
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(
    direction = "horizontal", title.position = "top")) +
  theme_bw() + theme(
    legend.key.height = unit(0.3, "cm"),
    legend.position = c(0.25,0.13),
    axis.title = element_blank())
```

\clearpage

### 3. Spatiotemporal random effects

@@

```{r plot-spatiotemporal-effects, echo=TRUE, eval=T}
ggplot(data = filter(p_proj, year > 1995)) +
  geom_point(data = filter(p, year > 1995), aes(X, Y, colour = epsilon_st), 
             size = 0.4, alpha = 0.5) +
  coord_sf(xlim = c(min(p_proj$X)-50000, max(p_proj$X)-50000), 
           ylim = c(min(p_proj$Y), max(p_proj$Y))) +
  scale_colour_viridis_c(guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

\clearpage

# References

::: {#refs}
:::
